Evolutionary rescue

### 2022-10-02 EvolutionaryRescue1.tex
Add files via upload

\usetikzlibrary{shadings}
\usetikzlibrary{backgrounds}
\usetikzlibrary{calc}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.text}

@@

\begin{document}
\title{Evolutionary rescue}
\author[1]{Remus Stana}
\author[1]{Yoav Ram}
\author[2]{‪Daniel Weissman}
\affil[1]{School of Zoology, Tel Aviv University, Tel Aviv-Yafo, Israel}
\affil[2]{Department of Physics, Emory University, Atlanta, United States of America}
\maketitle
\begin{abstract}
\end{abstract}

@@

\section{Model}
We model evolutionary rescue of cancer cells through aneuploidy with the help of multitype branching processes: we have three populations of cancer cells, wildtype, aneuploidy and mutant cells which proliferate and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively. Wildtype cells can acquire aneuploidy at rate $u$ while aneuploidy and wildtype cells can mutate with rate $v$ (se Figure \ref{figureAneuploidy}). The probability that a single mutant cell will survive is given by\cite{allen2010introduction}:
\begin{equation}
p_m=\left\{
  \begin{array}{@{}ll@{}}
  \frac{\lambda_m-\mu_m}{\lambda_m},\quad\text{if }\lambda_m>\mu_m \\
   0,\quad\text{else}.
  \end{array}\right.
  \end{equation}
 If the population originally consists of a single aneuploidy cell, then the probability that the population will survive is given by the following quadratic equation:
\begin{align}
1-p_a=\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2,
\end{align}
whose smallest solution is the survival probability:
\begin{align}\label{survproba}
p_a=\frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}.
\end{align}
Analogously, the survival probability of a population consisting initially of one wildtype cell satisfies the quadratic equation:
\begin{align}\nonumber
1-p_w&=\frac{\mu_w}{\lambda_w+\mu_w+u+v}+\frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right)+\frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2\\
&+\frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right),
\end{align}
where the smallest solution gives the survival probability:
\begin{align}\label{survprobw}
p_w=\frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w}.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\begin{tikzpicture}[very thick,decoration={
    markings,
    mark=at position 0.5 with {\arrow{>}}}, middlearrow/.style 2 args={
        decoration={             
            markings, 
            mark=at position 0.5 with {\arrow[xshift=3.333pt]{triangle 45}, \node[#1] {#2};}
        },
        postaction={decorate}
    }]
\edef\x{6.5} 
\edef\y{-0.5} 
\edef\r{1.25}

@@

\edef\y{-0.25}

\node at (\y,0) {Wildtype};
\node at (\y+5,0) {Aneuploidy};
\node at (\y+10,0) {Mutant};
\end{tikzpicture}
\caption{Diagram of the evolutionary rescue model where the population of cancer cells is subdivided into three populations of cancer cells, wildtype, aneuploidy and mutant cells which proliferate and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively. Wildtype cells can acquire aneuploidy at rate $u$ while aneuploidy and wildtype cells can mutate with rate $v$.}
\label{figureAneuploidy}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\usetikzlibrary{shadings}
\usetikzlibrary{backgrounds}
\usetikzlibrary{calc}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.text}
\usetikzlibrary{shapes.multipart}

@@

\begin{document}
\title{Evolutionary rescue}
\author[1]{Remus Stana}
\author[2]{‪Daniel Weissman}
\author[1]{Yoav Ram}
\affil[1]{School of Zoology, Tel Aviv University, Tel Aviv-Yafo, Israel}
\affil[2]{Department of Physics, Emory University, Atlanta, United States of America}
\maketitle
\begin{abstract}
Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction. A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments. We are interested in the mechanisms through which a population of cancer cells are able to adapt to chemotherapy, and in particular, the role played by chromosomal instability (aneuploidy). Cancer cells which have aneuploidy are hypothesized to have a higher fitness in an environment altered by anti-cancer drugs as they have incomplete pathways which drugs activate in order to kill the cells. Aneuploidy is highly prevalent in tumors and certain drugs which attempt to combat cancers through increasing chromosomal instability. As a result, the question we wish to answer is how aneuploidy impacts the fate of the population of cancer cells. We propose to model evolutionary rescue with the help of multi-type branching processes to obtain the probability that cancer will survive. Additionally, we will utilize large genomic datasets to asses the effects of aneuploidy on the probability of evolutionary rescue.
\end{abstract}

@@

\section{Model}
We model evolutionary rescue of cancer cells through aneuploidy with the help of multitype branching processes: we have three populations of cancer cells, wildtype euploid $w_t$, wildtype aneuploid $a_t$ and mutant euploid cells $m_t$ which proliferate and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively. Wildtype cells can acquire aneuploidy at rate $u$ while aneuploidy and wildtype cells can mutate with rate $v$ (se Figure \ref{figureAneuploidy}). Changes in the number of wildtype euploid, wildtype aneuploid and mutant euploid cells occur at rates:
\begin{align*}
w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
w_t&\rightarrow w_t-1:\quad \left(\mu_w+u\right)w_t,\\
a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
a_t&\rightarrow a_t-1:\quad \left(\mu_a-w\right)a_t,\\
m_t&\rightarrow m_t+1:\quad \lambda_am_t+wa_t,\\
m_t&\rightarrow m_t-1:\quad \mu_am_t.
\end{align*}
The survival probability of a population consisting initially of one wildtype cell satisfies the quadratic equation:
\begin{align}\nonumber
1-p_w&=\frac{\mu_w}{\lambda_w+\mu_w+u+v}+\frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right)+\frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2\\ \label{quadraticeqev1}
&+\frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right),
\end{align}
where the $p_a$ and $p_w$ are the survival probabilities of a population consisting initially of one aneuploidy and wildtype cell, respectively.

 The smallest solution of the quadratic equation \eqref{quadraticeqev1} gives the survival probability:
\begin{align}\label{survprobw}
p_w=\frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w}.
\end{align}
If the population originally consists of a single aneuploidy cell, then the probability that the population will survive is given by the following quadratic equation:
\begin{align}
1-p_a=\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2,
\end{align}
whose smallest solution is the survival probability:
\begin{align}\label{survproba}
p_a=\frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}.
\end{align}
The probability that a single mutant cell will survive is given by \cite{allen2010introduction}:
\begin{equation}
p_m=\left\{
  \begin{array}{@{}ll@{}}
  \frac{\lambda_m-\mu_m}{\lambda_m},\quad\text{if }\lambda_m>\mu_m \\
   0,\quad\text{else}.
  \end{array}\right.
  \end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\begin{tikzpicture}[very thick,decoration={
    markings,
    mark=at position 0.5 with {\arrow{>}}}, middlearrow/.style 2 args={
        decoration={             
            markings, 
            mark=at position 0.5 with {\arrow[xshift=3.333pt]{triangle 45}, \node[#1] {#2};}
        },
        postaction={decorate}
    }]
\edef\x{6.5} 
\edef\y{-0.5} 
\edef\r{1.25}

@@

\edef\x{0.25} 
\edef\y{-0.25}

\node[text width=2cm,align=justify] at (\y,\x)  
{\begin{center}
Wildtype  euploid
\end{center}};

\node[text width=2cm,align=justify] at (\y+5,\x)  
{\begin{center}
Wildtype  aneuploid
\end{center}};

\node[text width=2cm,align=justify] at (\y+10,\x)  
{\begin{center}
Mutant  euploid
\end{center}};
\end{tikzpicture}
\caption{Diagram of the evolutionary rescue model where the population of cancer cells is subdivided to wildtype euploid, wildtype aneuploid and mutant euploid cells that divide and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively. Wildtype cells can acquire aneuploidy at rate $u$. Aneuploid and wildtype cells can mutate with rate $v$.}
\label{figureAneuploidy}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
